---
title: "WGCNA_tutorial"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

We will use two datasets for this tutorial. The first is from Warner et al. 2018 and is from a Nematostella vectensis embryonic timeseries. The sampling is sparse, every 24 hours for 7 days, in duplicate (16 samples total). The second dataset is from Fisher et al. 2014 and is also from Nematostella vectensis. This timeseries is much denser, sampled hourly from 0-20 hours post fertilization in duplicate. For this tutorial we will subsample the data to include just 12-20 hours. The reason for this subsampling is to make the comparisons between the two datasets more straightforward and focus on zygotic transcription (assuming an MTZ of 10-12hpf).

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
